---
title: "The highest car prices in the world?"
author: Kenneth Lim
date: '2017-12-29'
slug: car-prices
categories: 
  - singapore
tags: 
  - cars
  - transport
output:
  blogdown::html_page:
    toc: false
    number_sections: no
    fig_width: 8
    fig_height: 8
---
```{r setup1, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(scipen = 999)

#load required packages
library(tidyverse); library(readr); library(lubridate)

#read data set
df <- read_csv("~/Documents/backup/R/data_gov_sg/coe-bidding-results/coe-results.csv")

#Tidy up the data set
df <- df %>%
  rename(year_month = month) %>%
  mutate(bidding_no = as.factor(bidding_no),
         date = as.Date(paste(year_month, "-01", sep = "")),
         month = month(date, label = TRUE),
         year = year(date),
         no_days = days_in_month(date),
         vehicle_class = as.factor(vehicle_class)) %>%
  rename(premium_new = premium,
         category = vehicle_class) %>%
  select(year_month, date, year, month, no_days, everything())

#Create a new data set that contains only COE for cars (Category D is for motorcycles)
df_cars <- filter(df, category != "Category D")

#Data checking
sum(df$bids_received <= df$bids_success) #1 instance
sum(df$quota > df$bids_received) #1 instance
sum(df$quota < df$bids_success)  #0

#Create a label we will use later
##Names taken from: https://www.mytransport.sg/oneMotoring/coeDetails.html
category_names <- c(
  `Category A` = "Category A: Car up to 1600cc & 97kW",
  `Category B` = "Category B: Car above 1600cc & 97kW",
  `Category C` = "Category C: Goods vehicle & bus",
  `Category D` = "Category D: Motorcycle", 
  `Category E` = "Category E: Open - all (excludes motorcycle from 2017)"
)

bidding_names <- c(
  `1` = "Bidding round 1",
  `2` = "Bidding round 2"
)
```

[Singapore is the most expensive place in the world to own a car, according to a survey by the Economist Intelligence Unit](http://www.todayonline.com/singapore/singapore-ranked-worlds-most-expensive-city-again). Currently, buying a new Toyota Corolla would cost around SGD \$100,000 (~USD \$75,000), and you can only own it for 10 years. After 10 years, you have to deregister and dispose your car or buy a Certificate of Entitlement to continue using it for another 5 or 10 years.

According to the _Land Transport Authority of Singapore_^[https://www.lta.gov.sg/content/ltaweb/en/roads-and-motoring/owning-a-vehicle/vehicle-quota-system/certificate-of-entitlement-coe.html],

> Anyone who wishes to register a new vehicle in Singapore must first obtain a Certificate of Entitlement (COE), in the appropriate vehicle category. A COE represents a right to vehicle ownership and use of the limited road space for 10 years... COEs are released through competitive bidding. There are 2 bidding exercises every month. 

When buying a new car, you have to obtain a COE _and_ pay for the cost of the car. COEs are used together with a quota on the number of vehicles allowed on the roads to regulate the vehicle population growth^[for more details: http://eresources.nlb.gov.sg/infopedia/articles/SIP_1005_2006-04-07.html].

Using data from the Singapore government's open-data initiative^[https://data.gov.sg/dataset/coe-bidding-results], we can find out:

+ How much does a COE for a new car cost?

+ How many COEs are available?

The charts below show the cost of COEs for new cars in each bidding round every month, from January 2010 to August 2017. Each panel represents a different vehicle category.

```{r prices}

df_cars %>%
  ggplot(mapping = aes(x = date, y = premium_new, col = bidding_no)) + 
  geom_line(size = 0.7) +
  geom_point(size = 0.8, col = "white") +
  geom_point(size = 0.45, alpha = 0.85) +
  scale_x_date(breaks = seq(as.Date("2010-01-01"), as.Date("2017-08-01"), by = "6 months"),
               limits = c(as.Date("2010-01-01"), as.Date("2017-08-01")),
               date_labels= "%b'%y") +
  scale_y_continuous(breaks = seq.int(from = 20000, to = 100000, by = 10000)) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8.5),
        axis.text.y = element_text(size = 8.5),
        axis.title = element_text(size = 9),
        title = element_text(size = 10),
        strip.text = element_text(size = 8),
        legend.text = element_text(size = 8),
        panel.grid.minor.x = element_blank(),
        legend.position = "bottom",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-17,0,2,0)) + 
  facet_wrap( ~ category, labeller = as_labeller(category_names), ncol = 2) + 
  labs(x = "",
       y = "Cost of a COE for a new car (SGD $)",
       col = "Bidding round",
       title = "COE premiums for a new car, Jan 2010--Aug 2017",
       subtitle = "Data from data.gov.sg")

```

The cost of COEs tended to increase initially and then decrease after 2013, although there were were rather large fluctuations. The fluctations tended to be smaller after 2016. 

The cost of COEs tended to be similar across each bidding round, with some notable exceptions. For example, see Category B just after January 2016, where the COE price in round 1 was noticeably lower than in round 2.

Categories B and E tended to have very similar patterns. The price of COEs for new Category B and new Category E cars tended to be very similar in each month.

The charts below show the number of available COEs for cars in each bidding round every month, from January 2010 to August 2017. Each panel represents a different vehicle category.

```{r quota}

df_cars %>%
  ggplot(mapping = aes(x = date, y = quota, col = bidding_no)) + 
  geom_line(size = 0.7) +
  geom_point(size = 0.7, col = "white") +
  geom_point(size = 0.4, alpha = 0.85) +
  scale_x_date(breaks = seq(as.Date("2010-01-01"), as.Date("2017-08-01"), by = "6 months"),
               limits = c(as.Date("2010-01-01"), as.Date("2017-08-01")),
               date_labels= "%b'%y") +
  scale_y_continuous(breaks = seq.int(from = 200, to = 2200, by = 200)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8.5),
        axis.text.y = element_text(size = 8.5),
        axis.title = element_text(size = 9),
        title = element_text(size = 10),
        strip.text = element_text(size = 8),
        legend.text = element_text(size = 8),
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-17,0,2,0)) + 
  facet_wrap( ~ category, labeller = as_labeller(category_names), ncol = 2) + 
  labs(x = "",
       y = "Number of COEs available",
       col = "Bidding round",
       title = "COE quota for cars, Jan 2010--Aug 2017",
       subtitle = "Data from data.gov.sg")
```

For each category, the number of available COEs tended to be very similar in each bidding round, with a few exceptions. For example, see the red spike in Category E, between January and July 2016.

Across categories, the trend in the number of available COEs was different. 

There is an interesting pattern in Category C, from 2014 onwards. There is a jump from April to May. From May to July, it stays constant. Then, it falls in August. This pattern is most noticeable in 2017.
