---
title: Are there more males or females in Singapore?
author: Kenneth Lim
date: '2017-12-10'
slug: singapore-sex-ratio
categories:
  - singapore
tags:
  - population
  - gender
  - ethnicity
draft: no
output:
  blogdown::html_page:
    toc: false
    number_sections: no
    fig_width: 6
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(scipen = 999)
library(readxl); library(tidyverse); library(forcats)
#Read dataset
df <- read_excel("~/Documents/backup/R/data_gov_sg/sg_stats_official_2017.xlsx", 
                         sheet = "T2", na = "na") 

df <- df[c(5, 6, 29, 52, 75, 98, 121, 144, 167, 190, 213, 236, 259, 282, 305, 328), ] #extract specific rows
colnames(df) <- df[1, ] #make first row the header
df <- t(df) #transpose df
rownames(df) <- NULL #remove rownames
colnames(df) <- df[1, ] #make first row the header
df <- df[-1,] # remove first row
df <- gsub(',', '', df) #remove commas in values
df <- as_tibble(df) #make data frame a tibble

df <- data.frame(lapply(df, function(x) as.integer(x))) #convert all variables from character to integer type

df <- df %>% #rename variables
  rename(year = "Variables",
         residents_all = "Total.Residents",
         residents_m = "Total.Male.Residents",
         residents_f = "Total.Female.Residents",
         malay_all = "Total.Malays",
         malay_m = "Total.Male.Malays",
         malay_f = "Total.Female.Malays",
         chi_all = "Total.Chinese",
         chi_m = "Total.Male.Chinese",
         chi_f = "Total.Female.Chinese",
         ind_all = "Total.Indians",
         ind_m = "Total.Male.Indians",
         ind_f = "Total.Female.Indians",
         other_all = "Other.Ethnic.Groups..Total.",
         other_m = "Other.Ethnic.Groups..Males.",
         other_f = "Other.Ethnic.Groups..Females.") 

y_skip <- rep("" , times = 4)
year_label <- c("1960", y_skip, "1965", y_skip, "1970", y_skip, "1975", y_skip, "1980", y_skip, 
           "1985", y_skip, "1990", y_skip, "1995", y_skip, "2000", y_skip, "2005", y_skip, 
           "2010", y_skip, "2015", "", "2017")
```

Try to guess. Are there more males or females in Singapore? What about across each ethnic group? Read on and compare your answers to the data provided by the Singapore Department of Statistics!

```{r sexdiff, include = FALSE}
df_grow <- df %>%
  mutate(f_grow = residents_f - lag(residents_f),
         m_grow = residents_m - lag(residents_m),
         difference = residents_f - residents_m)

df_residents <- df %>% 
  select(year, residents_m, residents_f) %>%
  rename("Male" = "residents_m", 
         "Female" = "residents_f") %>%
  gather(key = sex, value = number, Male:Female, factor_key = TRUE)

df_sr <- df %>% #residents sex ratio
  mutate(residents_sr = (residents_m / residents_f) * 100)

df_ethnic <- df %>% #residents sex ratio by ethnic group
  mutate(Chinese = (chi_m/chi_f)*100,
         Malay = (malay_m/malay_f)*100,
         Indian = (ind_m/ind_f)*100,
         Other = (other_m/other_f)*100) %>%
  select(year, Chinese:Other) %>%
  gather(key = Ethnicity, value = sex_ratio, Chinese:Other, factor_key = TRUE)

df_1 <- df %>%
  mutate(chi_sr = (chi_m/chi_f)*100,
         mly_sr = (malay_m/malay_f)*100,
         ind_sr = (ind_m/ind_f)*100,
         oth_sr = (other_m/other_f)*100)

df_chg <- df_1 %>%
  mutate(chi_chg = chi_sr - lag(chi_sr),
         mly_chg = mly_sr - lag(mly_sr),
         ind_chg = ind_sr - lag(ind_sr),
         oth_chg = oth_sr - lag(oth_sr)) 
  
df_mlychg <- df_chg %>%
  select(year, mly_sr, mly_chg) %>%
  filter(mly_chg >= 0)
```

For all charts below, data before 1980 (except 1970) refer to the 'total population' (citizens, permanent residents, and non-residents), while data for 1970 and 1980 onwards refer to 'residents' (citizens and permanent residents).^[According to www.singstat.gov.sg: 'Data for 1970 and 1980 refer to all residents present in Singapore on Census day.  Data from 2000 onwards are based on the register-based approach.  Data from 2003 onwards exclude residents who are overseas for a continuous period of 12 months or longer as at the reference period.']

**The chart below shows the number of males and females in Singapore, for each year from 1960 to 2017.**

```{r numbersex}
df_residents %>%
  ggplot(mapping = aes(x = year, y = number/10^6, col = fct_rev(sex))) + 
  geom_vline(xintercept = c(1970, 1980, 2000, 2003), linetype = 2, alpha = 0.25) +
  geom_line(size = 1) +
  geom_point(size = 1.75, col = "white") +
  geom_point(size = 1) +
  scale_y_continuous(breaks = seq.int(from = 0.8, to = 2.1, by = 0.1), limits = c(0.75, 2.1)) +
  scale_x_continuous(breaks = seq.int(from = 1960, to = 2017, by = 1), labels = year_label) +
  theme_bw() + 
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "bottom",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-7, 0, 2, 0),
        plot.caption = element_text(size = 7.5)) +
  labs(x = "Year",
       y = "Number of people (Million)",
       col = "",
       title = "Number of males and females in Singapore, 1960--2017",
       subtitle = "Data from www.singstat.gov.sg",
       caption = "Data before 1980 (except 1970) refer to: 'Total population'.\n Data for 1970 and from 1980 refer to: 'Residents'.\n 1970, 1980 and from 2000 and 2003 onwards: Change in how residents were counted.")
```

+ There were more males from `r min(df_grow$year[df_grow$difference < 0])` to `r max(df_grow$year[df_grow$difference < 0])`. From `r min(df_grow$year[df_grow$difference > 0])` to `r max(df_grow$year[df_grow$difference > 0])`, there were more females.

+ The number of females increased by an average (median) of `r format(signif(median(df_grow$f_grow, na.rm = TRUE), digits = 3), big.mark = ",")` each year.

+ The number of males increased by an average (median) of `r format(signif(median(df_grow$m_grow, na.rm = TRUE), digits = 3), big.mark = ",")` each year.

+ There were only `r nrow(filter(df_grow, f_grow < 0 | m_grow < 0))` years where there was a drop in the number of males and females: `r unique(df_grow$year[(df_grow$f_grow <0 | df_grow$m_grow < 0) & df_grow$year != 1960])`. This is likely due to the change in how residents were counted. For example, data for 1970 and 1980 refer to all residents present in Singapore on Census day. 
    
    + The biggest drop happened in 1980: there were `r format(signif(abs(min(df_grow$m_grow[df_grow$year != 1960])), digits = 3), big.mark = ",")` fewer males and `r format(signif(abs(min(df_grow$f_grow[df_grow$year != 1960])), digits = 3), big.mark = ",")` fewer females compared to the previous year, 1979. 

**The chart below shows the sex ratio of Singapore for each year from 1960 to 2017.**

The sex ratio is the number of males for every 100 females. A sex ratio of 100 means there are 100 males for every 100 females (an equal number of males and females). A ratio above 100 means there are more males. A ratio below 100 means there are more females.

```{r sexratio1}
df %>%
  mutate(residents_sr = (residents_m / residents_f) * 100) %>%
  ggplot(mapping = aes(x = year, y = residents_sr)) + 
  geom_hline(yintercept = 100, col = "grey80") + 
  geom_line(size = 1) +
  geom_point(size = 1.75, colour = "white") + 
  geom_point(size = 1) + 
  geom_vline(xintercept = c(1970, 1980, 2000, 2003), linetype = 2, alpha = 0.3) + 
  scale_x_continuous(breaks = seq.int(from = 1960, to = 2017, by = 1), labels = year_label) +
  scale_y_continuous(breaks = seq.int(from = 95, to = 110, by = 1), limits = c(95,110)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1),
        plot.caption = element_text(size = 7.5)) +
  labs(x = "Year\n",
       y = "Number of males for every 100 females",
       title = "Singapore's sex ratio, 1960--2017",
       subtitle = "Data from www.singstat.gov.sg",
       caption = "Data before 1980 (except 1970) refer to: 'Total population'.\n Data for 1970 and from 1980 refer to: 'Residents'.\n 1970, 1980 and from 2000 and 2003 onwards: Change in how residents were counted.")

```

+ There were around `r round(df_sr$residents_sr[df_sr$year == min(df_sr$year)])` males for every 100 females in `r min(df_sr$year)`. The sex ratio tended to decrease each year to around `r round(df_sr$residents_sr[df_sr$year == max(df_sr$year)])` males for every 100 females in `r max(df_sr$year)`.^[The sex ratios were rounded to the nearest whole number.]

**The chart below shows the sex ratio for each ethnic group, for each year from 1960 to 2017.**

```{r sexratio2}
df_ethnic %>%
  ggplot(mapping = aes(x = year, y = sex_ratio, col = fct_reorder2(Ethnicity, year, sex_ratio))) + 
  geom_hline(yintercept = 100, col = "grey80") + 
  geom_line(size = 1) +
  geom_point(size = 1.75, colour = "white") +
  geom_point(size = 1) +
  geom_vline(xintercept = c(1970, 1980, 2000, 2003), linetype = 2, alpha = 0.3) + 
  scale_x_continuous(breaks = seq.int(from = 1960, to = 2017, by = 1), labels = year_label) +
  scale_y_continuous(breaks = seq.int(from = 80, to = 200, by = 10), limits = c(80, 200)) +
  theme_bw() + 
    theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "bottom",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(-7, 0, 2, 0),
        plot.caption = element_text(size = 7.5)) +
  labs(x = "Year",
       y = "Number of males for every 100 females",
       col = "Ethnic group",
       title = "Sex ratio for each ethnic group in Singapore, 1960--2017",
       subtitle = "Data from www.singstat.gov.sg",
       caption = "Data before 1980 (except 1970) refer to: 'Total population'.\n Data for 1970 and from 1980 refer to: 'Residents'.\n 1970, 1980 and from 2000 and 2003 onwards: Change in how residents were counted.\n From 1970, Indians include Sri Lankans (Ceylonese) who were previously classified under 'Others'.")
  
```

The sex ratio within each ethnic group tended to decrease over the years.^[The sex ratios quoted below for each ethnic group were rounded to the nearest whole number.]

+ **Indians.** In `r min(df_1$year)`, there were `r round(df_1$ind_sr[df_1$year == min(df_1$year)])` Indian males for every 100 Indian females. In `r max(df_1$year)`, there were `r round(df_1$ind_sr[df_1$year == max(df_1$year)])` Indian males for every 100 Indian females.

+ **Malays.** In `r min(df_1$year)`, there were `r round(df_1$mly_sr[df_1$year == min(df_1$year)])` Malay males for every 100 Malay females. In `r max(df_1$year)`, there were `r round(df_1$mly_sr[df_1$year == max(df_1$year)])` Malay males for every 100 Malay females. There was a visible increase in the sex ratio from `r min(df_mlychg$year)` to `r df_mlychg$year[df_mlychg$year==1979]`.

+ **Chinese.** In `r min(df_1$year)`, there were `r round(df_1$chi_sr[df_1$year == min(df_1$year)])` Chinese males for every 100 Chinese females. In `r max(df_1$year)`, there were  `r round(df_1$chi_sr[df_1$year == max(df_1$year)])` Chinese males for every 100 Chinese females.

+ **Others.** In `r min(df_1$year)`, there were `r round(df_1$oth_sr[df_1$year == min(df_1$year)])` "Other" males for every 100 "Other" females. In `r max(df_1$year)`, there were  `r round(df_1$oth_sr[df_1$year == max(df_1$year)])` "Other" males for every 100 "Other" females. The sex ratio was highest in `r df_1$year[df_1$oth_sr == max(df_1$oth_sr)]` (`r round(max(df_1$oth_sr))` "Other" males for every 100 "Other" females).

**What could explain the change in Singapore's sex ratio, especially in the last 10 years or so?** Immigration could be one reason, according to an article in the Straits Times newspaper.^[Rachel Lin. "Influx of women tilts Singpore's sex ratio." _The Straits Times_ 04 September 2010. Accessed on 10 Dec 2017 from: http://lkyspp2.nus.edu.sg/ips/wp-content/uploads/sites/2/2013/06/ST_Influx-of-women-tilts-Singapore_040910.pdf.] [Click here to read the Straits Times news article (pdf).](http://lkyspp2.nus.edu.sg/ips/wp-content/uploads/sites/2/2013/06/ST_Influx-of-women-tilts-Singapore_040910.pdf)