---
title: 'Hospital Bed Numbers: How does Singapore compare with the OECD?'
author: Kenneth Lim
date: '2019-03-28'
slug: hospital-beds-in-singapore
categories:
  - singapore
  - health
tags:
  - hospital
  - population
output:
  blogdown::html_page:
    toc: false
    number_sections: no
    fig_width: 7    
---
```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(scipen = 999999)

###Cleaning, transforming, and tidying data for analysis

#Load packages
library(readr); library(readxl); library(ggplot2); library(dplyr); library(tidyr); library(blogdown); library(ggrepel)
theme_set(theme_bw())

#Load 2 datasets for the number of hospital beds and population statistics
df_beds <- read_csv("~/Documents/backup/R/data_gov_sg/number-of-hospital-beds/number-of-hospital-beds-annual.csv") #name dataset df_beds
df_sgstats <- read_excel("~/Documents/backup/R/data_gov_sg/sg_stats_official.xlsx", 
                         sheet = "T1", skip = 2) #name dataset df_sgstats

#Tidying the population statistics dataset with some transformations
x1 <- df_sgstats[3,-1] #select first row (year), without first column
x1 <- as.integer(x1) #make into integers

x2 <- df_sgstats[4,-1] #select second row (population), without first column
x2 <- gsub(',', '', x2) #remove commas from numbers
x2 <- as.integer(x2) #make into integers

#Creating a new population statistics data frame that is tidied
df_allpop <- data.frame(x1, x2) %>%
  rename(year = x1, total_population = x2) %>%
  filter(year < 2015) #time frame of the hospital beds dataset

df <- df_beds %>%
  tidyr::spread(sector, no_of_hospital_beds) %>% #Make into two columns 
  rename(private = `Private Sector`, #Rename columns
         public = `Public Sector`) %>%
  mutate(all_beds = private + public) %>%
  right_join(df_allpop, by = "year")
```

In 2014, the main Singapore newspaper reported that public hospitals were facing a 'severe bed crunch' and putting patients in 'beds along corridors, with just curtains to give them privacy' when demand spikes.^[http://www.straitstimes.com/singapore/hospitals-facing-severe-bed-crunch-take-unusual-steps] ^[http://www.straitstimes.com/singapore/health/long-waits-at-aes-despite-more-beds] Some citizens have voiced concerns about the number of hospital beds^[https://www.theonlinecitizen.com/2010/11/22/not-so-ordinary-ordinary-singaporean-leong-sze-hian/].

In this post, you will see how the number of hospital beds has _increased_, while the number of hospital beds per 1,000 people has _decreased_, over the years. ^[Data available from 1960 to 2014: https://data.gov.sg/dataset/21d426ab-3a21-46ae-8602-99a5a1968593/download, accessed 06 Feb 2017] The downward trend in the number of hospital beds per 1,000 people in Singapore is consistent with most OECD countries. Some possible reasons for this trend are given at the end.


## Total number of hospital beds in Singapore
The total number of hospital beds in Singapore increased from `r formatC(df$all_beds[df$year==1960], big.mark = ",")` in 1960 to `r formatC(df$all_beds[df$year==2014], big.mark = ",", format = "d")` in 2014, which is a `r round(((df$all_beds[df$year==2014] - df$all_beds[df$year==1960])/df$all_beds[df$year==1960])*100, digits = 1)`% change. The chart below shows the number of private and public hospital beds for each year from 1960 to 2014^[From the metadata provided: 'Public sector hospitals include government hospitals, government restructured hospitals and specialty centres'. 'Private sector hospitals include community hospitals and chronic sick hospitals.'].

```{r, fig.align='center'}
blank_9 <- rep("", 9)
year_lab <- c("1960", blank_9, "1970", blank_9, "1980", blank_9, "1990", blank_9, "2000", blank_9, "2010", rep("", 3), "2014")

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
cols <- gg_color_hue(2)

df %>%
  select(year:public) %>%
  tidyr::gather(type, number, public:private) %>%
  ggplot(aes(x = year, y = number, col = type, linetype = type)) +
  geom_line(size = 0.85) +
  geom_vline(xintercept = c(1960, 1970, 1980, 1990, 2000, 2010),
             alpha = 0.05) +
  scale_y_continuous(breaks = seq(from = 0, to = 10000, by = 1000),
                     labels = c(0:10)) +
  scale_x_continuous(breaks = seq(from = 1960, to = 2014, by = 1),
                     labels = year_lab) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(size = 14),
        legend.position = "none") +
  labs(x = "Year",
       y = "Number of hospital beds (Thousand)",
       title = "Number of private and public hospital beds in Singapore, 1960--2014",
       caption = "chart: numbers.sg\n data source: data.gov.sg") +
  geom_label(aes(x = 2014, y = 2000), size = 3.2, label = "Private", col = cols[1]) +
  geom_label(aes(x = 2014, y = 8500), size = 3.2, label = "Public", col = cols[2])
```

+ Public hospital beds increased from `r formatC(df$public[df$year == 1960], big.mark = ",")` in 1960 to `r formatC(df$public[df$year == 2014], big.mark = ",")` in 2014, which is an increase of `r round(((df$public[df$year == 2014] - df$public[df$year == 1960])/df$public[df$year == 1960])*100, digits = 0)`%.

+ Private hospital beds increased from `r formatC(df$private[df$year == 1960], big.mark = ",")` in 1960 to `r formatC(df$private[df$year == 2014], big.mark = ",")` in 2014, which is an increase of `r round(((df$private[df$year == 2014] - df$private[df$year == 1960])/df$private[df$year == 1960])*100, digits = 0)`%.

## Hospital beds per 1,000 people

```{r }
df_1k <- df %>%
  mutate(all_1k = (all_beds/total_population)*1000,
         all_1k2 = all_beds * (total_population/1000))
```

Singapore's total population has changed over the years ([see previous blog post by clicking here](https://numbers.sg/posts/the-number-of-citizens-permanent-residents-and-non-residents-in-singapore/)). We can take this into consideration by calculating how many hospital beds there are, on average, for every 1,000 people in Singapore. The chart below shows the number of private and public hospital beds per 1,000 people from `r min(df_1k$year)` to `r max(df_1k$year)`. 

```{r}
df_1k <- df %>%
  select(-all_beds) %>%
  tidyr::gather(type, number, public:private) %>%
  mutate(per_1k = (number/total_population)*1000) 
```

```{r, fig.align='center'}
df_1k %>%
  ggplot(aes(x = year, y = per_1k, col = type, linetype = type)) +
  geom_line(size = 0.85) +
  geom_vline(xintercept = c(1960, 1970, 1980, 1990, 2000, 2010),
             alpha = 0.05) +
  scale_y_continuous(breaks = seq(from = 0, to = 5, by = 0.5)) +
  scale_x_continuous(breaks = seq(from = 1960, to = 2014, by = 1),
                     labels = year_lab) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(size = 14),
        legend.position = "none") +
  labs(x = "Year",
       y = "Number of hospital beds per 1,000 people",
       title = "Number of hospital beds in Singapore per 1,000 people, 1960--2014",
       caption = "chart: numbers.sg\n data source: data.gov.sg, SingStat") +
  geom_label(aes(x = 2014, y = 0.75), size = 3.2, label = "Private", col = cols[1]) +
  geom_label(aes(x = 2014, y = 2), size = 3.2, label = "Public", col = cols[2])
```

+ The number of public hospital beds per 1,000 people declined from `r round(df_1k$per_1k[df_1k$year==min(df_1k$year) & df_1k$type == "public"], digits = 1)` in `r min(df_1k$year)` to `r round(df_1k$per_1k[df_1k$year==max(df_1k$year) & df_1k$type == "public"], digits = 1)` in `r max(df_1k$year)`. 

+ There is a slight increase in the number of private hospital beds per 1,000 people from `r round(df_1k$per_1k[df_1k$year==min(df_1k$year) & df_1k$type == "private"], digits = 1)` in `r min(df_1k$year)` to `r round(df_1k$per_1k[df_1k$year==max(df_1k$year) & df_1k$type == "private"], digits = 1)` in `r max(df_1k$year)`. 

## OECD comparison: hospital beds per 1,000 people

```{r}
sg <- df %>%
  mutate(country = "SGP",
         per_1k = (all_beds/total_population)*1000) %>%
  select(-c(private, public, total_population, all_beds)) %>%
  tidyr::gather(x, per_1k, per_1k) %>%
  select(-x)

oecd <- read.csv("~/Documents/backup/R/data_gov_sg/number-of-hospital-beds/oecd_hospital_beds_annual.csv", stringsAsFactors=FALSE)

oecd <- oecd %>%
  rename(year = TIME,
         country = LOCATION,
         per_1k = Value) %>%
  select(year, country, per_1k) 
  
oecd_other <- oecd %>%
  filter(country == "CHN" | country == "KOR" | country == "JPN" | country == "GBR" | country == "IND" | country == "USA") %>%
  rbind(sg)
```

The chart below shows how Singapore (SGP) compares with `r length(unique(oecd$country))` OECD member and key partner countries^[list of countries and OECD data: https://data.oecd.org/healtheqt/hospital-beds.htm, accessed 27 March 2019]. To make the chart easier to read, only a few countries are labelled and emphasised^[JPN: Japan, KOR: Korea, CHN: China, USA: United States of America, GBR: United Kingdom, SGP: Singapore, IND: India].

```{r, fig.align='center'}
oecd %>%
  filter(country != "USA" & country != "GBR" & country != "JPN" & country != "KOR" & country != "CHN" & country != "IND") %>%
  ggplot(aes(x = year, y = per_1k, group = country)) +
  geom_line(alpha = 0.2, size = 0.3) +
  geom_line(data = oecd_other, col = "black", alpha = 0.5, size = 0.5) +
  geom_line(data = sg, col = "red", size = 0.8) +
  ggrepel::geom_label_repel(data = filter(oecd_other, year == 2014), 
                            aes(label = country), size = 3, direction = "y", 
                            hjust = -0.75, seed = 2, segment.size = 0.4,
                            min.segment.length = 0) +
  scale_x_continuous(breaks = seq(from = 1960, to = 2014, by = 1),
                     labels = year_lab, limits = c(1960, 2015.5)) +
  scale_y_continuous(breaks = seq(from = 0, to = 16, by = 2)) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(size = 14)) +
  labs(x = "Year",
       y = "Number of hospital beds per 1,000 people",
       title = "SGP vs. OECD: Number of hospital beds per 1,000 people, 1960--2014",
       caption = "chart: numbers.sg\n data source: data.gov.sg, SingStat, OECD")
```

+ There is a downward trend in almost all countries, with some notable exceptions like Korea (KOR) and China (CHN).

+ In 2014, the median number of hospital beds per 1,000 people in the OECD countries was `r round(median(oecd$per_1k[oecd$year == 2014]), digits = 1)`. In 2014, Singapore had `r round(sg$per_1k[sg$year == 2014], digits = 1)` hospital beds per 1,000 people.

## Reasons for decline in hospital beds per 1,000 people
There are many possible reasons for a decreasing number of hospital beds per 1,000 people. This includes having a population that is increasing faster than the number of hospital beds being installed, as well as better medical technology and more services at the community level, both of which reduce the need for hospitalisation^[https://fullfact.org/health/do-we-have-fewer-hospital-beds-most-europe/].

